---
title: "Chapter 19: QTL"
date: "11-25-2018"
author: "Jinliang Yang"
output:
   rmdformats::material:
      highlight: kate
      self_contained: no
      code_folding: hide
---

# Git version control

```{bash, eval=FALSE}
git clone git@github.com:jyanglab/agro931-project.git
```

### Fork the project

Fork the project on github to your own repo and git clone it to your local computer

### Create or edit files

Folders in a project serve for their own purposes. Read [here](https://jyanglab.com/2017-01-07-project/) to learn more about them.

### Commit your changes
Note commit with informative message. 

```{bash, eval=FALSE}
git add --all
git commit -m "to explore litature in qtl from 1960-2018"
```

### push to the branch

```{bash, eval=FALSE}
git push
```


# Search pubmed using key words

```{r search}
### source https://freshbiostats.wordpress.com/2013/12/03/analysis-of-pubmed-search-results-using-r/
library(RCurl)
library(RISmed)
query = "QTL"

res <- EUtilsSummary(query, type="esearch", db = "pubmed", mindate=1960, maxdate=2018, retmax=30000)
QueryCount(res)
```

### Pubs by Years
```{r years, eval=FALSE}

qtlpub <- EUtilsGet(res)
years <- YearPubmed(qtlpub)
qc <- as.data.frame(table(years))
write.table(qc, "qtl_pub_count.csv", sep=",", row.names=FALSE, quote=FALSE)
#save(file="qtlpub", list="qtlpub")
```

### Plot the number of pubs by year
```{r}
library(tidyverse)
qc <- read.csv("qtl_pub_count.csv")
theme_set(theme_grey(base_size = 18)) 
s <- ggplot(qc, aes(x=years, y=Freq)) + 
    #opts(axis.text.x=theme_text(angle=90)) +
    geom_bar(stat="identity") +
    labs(x="", y="# of pubs") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12))
s
```


## What are the journals?

```{r}
### get the count of NGS pub per journal:
journal <- MedlineTA(qtlpub)
qtl_journal_count <- as.data.frame(table(journal))
qtl_journal_count_top25 <- qtl_journal_count[order(-qtl_journal_count[,2]),][1:25,]

as_tibble(qtl_journal_count_top25)
```

## Top authors
```{r}
### Get authors
auths <- Author(qtlpub)
Last <- sapply(auths, function(x) paste(x$LastName, x$ForeName))
auths2 <- as.data.frame(sort(table(unlist(Last)), dec=TRUE))
auths2$author <- row.names(auths2)
names(auths2)[1] <- c("num")

df <- as_tibble(auths2)
head(df, 30)
```

## Top Affiliations

```{r}
### Get affiliations
aff <- Affiliation(qtlpub)
#Last <- sapply(auths, function(x) paste(x$LastName, x$ForeName))
aff2 <- as.data.frame(sort(table(unlist(aff)), dec=TRUE))
aff2$id <- row.names(aff2)
names(aff2)[1] <- c("number")

df2 <- as_tibble(aff2)
head(df2, 30)


```











