---
title: "Changes in allele frequency"
author: "Jinliang Yang"
date: "Jan. 29, 2024"
output:
  xaringan::moon_reader: 
    css: ["default", "../../asset/unl-150.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Factors lead  to changes of allele freq

### Systematic processes

Predictable both in amount and in direction

- __Migration__

- __Mutation__

- __Selection__

--

### Stochastic process

Arises in small populations from the effect of sampling

- Predictable in amount but not in direction

---

# Migration

- Rate of change depends on rate of migration and difference in allele frequency between original pop and immigrants

- In agriculture, can think of cross-breeding as having a similar effect as migration

--

### In mixed population after migration

- $m$ = proportion of new immigrants (each generation)
- $q_m$ = allele freq in immigrant population
- $q_0$ = allele freq in original pop

--

\begin{align*}
q_1 & = mq_m + (1-m)q_0 \\
 & = m(q_m - q_0) + q_0 \\
\end{align*}

--

Change in allele freq:

\begin{align*}
\Delta q & = q_1 - q_0 = m(q_m - q_0)
\end{align*}

---

# Mutation

- To affect allele frequencies, a mutation has to be maintained!

- New mutation = heterozygote (most commonly)
  - 50% chance of being passed on 
  - After $n$ generations, the probability of the allele remains

\begin{align*}
(1 - \frac{1}{2})^n
\end{align*}

--

```{r echo=FALSE, fig.height=5}
n <- 1:10
plot(x=n, y=(1-1/2)^n, type="l", lwd=3)
```

---

# Mutation

### What is the true rate of mutation ( $\mu$ )? 

- Depends on organism and locus

--

- Human:
  - Point mutation (2.5 to 1 $\times 10^{-8}$ mutations/locus/generation)
    > Neel et al 1986; Nachman and Crowell 2000

- Viruses
  - $1 \times 10^{-3}$ to $1 \times 10^{-5}$ per base/generation
  
  - SARS-CoV-2: spontaneous mutation rate of $1.3 \times 10^{-6}$ (the below study found spike protein accumulated five times more mutations)
   > Amicone et al 2022


---

# Recurrent Mutation

- Forward mutation rate __ $\mu$__ from $A_1$ $\rightarrow$ $A_2$
- Backward mutation rate __ $\nu$__ from $A_1$ $\leftarrow$ $A_2$

- Given the initial allele freq $p_0$ for $A_1$ and $q_0$ for $A_2$

--

- Change $\Delta q = \mu p_0 - \nu q_0$

--

In equilibrium,

\begin{align*}
&\Delta q = \mu p_0 - \nu q_0 = 0 \\
& \mu p_0 = \nu q_0 \\
\end{align*}

or 

\begin{align*}
& \mu (1- q) = \nu q \\
& q = \frac{\mu}{\nu + \mu} \\
\end{align*}

---

# Mutation

- #### Mutation alone produces slow change in allele frequency

- #### But, mutation may be important in some small pops (later)

- #### And, mutation is the ultimate source of genetic variation










