---
title: "Chapter 11: Selection"
date: "11-12-2018"
author: "Jinliang Yang"
output:
   rmdformats::material:
      highlight: kate
      self_contained: no
      code_folding: hide
---

# A chicken selection example:

In a chicken breeding program, the breeders aim to increase the 56-day body weight.
They plan to conduct 40 discrete generations of selection.

## Breeding scheme:

The base population consists of 100 animals, of which 50 are male and 50 are female. In each generation, the best 5 males are selected on the basis of their phenotypic value for a single polygenic trait (56-day body weight) and mated to the females (N=50) to produce 100 replacement animals. 

### Assumptions:

1. The quantitative trait, 56-day body weight, is modeled as being controlled by 1,000 genetic loci (or QTLs). 
2. These QTL are equally split across 10 chromosomes so that there are 100 QTLs per chromosome. 


# Simulation

### Simulate a founder population with genomic data

```{r, eval=TRUE}
# install.packages("AlphaSimR")
library(AlphaSimR)

# Creating Founder population
founderPop = runMacs(nInd=1000, nChr=10, segSites=5000)
```


### Simulate the phenotype

1. The quantitative trait, 56-day body weight, is modeled as being controlled by 1,000 genetic loci (or QTLs). 
2. These QTL are equally split across 10 chromosomes so that there are 100 QTLs per chromosome. 
3. h2=0.5

```{r, eval=TRUE}
# Setting Simulation Parameters
SP = SimParam$new(founderPop)
SP$addTraitA(nQtlPerChr=100)
SP$setGender("yes_sys")
SP$setVarE(h2=0.5)
```



# Conduct the selection

```{r, eval=TRUE}
# Modeling the Breeding Program
pop = newPop(founderPop)
genMean = meanG(pop)
for(generation in 1:40){
  pop = selectCross(pop=pop, nFemale=50, nMale=5, use="pheno", nCrosses=100)
  genMean = c(genMean, meanG(pop))
}
```

### Examining the Results

```{r, eval=TRUE}
plot(0:40, genMean, type="p", xlab="Generation", ylab="Mean Genetic Value")
```











