---
title: "Effectiveness of Selection"
author: "Jinliang Yang"
date: "Jan. 31, 2024"
output:
  xaringan::moon_reader: 
    css: ["default", "../../asset/unl-150.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```



# Selection against the recessive allele

- Gametic contribution is the product of genotype freq and fitness

|                    | $A_1A_1$   | $A_1A_2$   | $A_2A_2$   |
| :-------: | : ------ : | :-------: | :-------: |
| Initial frq           | $p^2$    | $2pq$    | $q^2$     | 
| Coefficient of selection | $0$   | $0$      | $s$       |
| Fitness ( $w$ )   |  $1$        |  $1$        |  $1-s$   | 
| Gametic contribution | $p^2$   | $2pq$    | $q^2(1-s)$    |

---

# Selection against the recessive allele

- Gametic contribution is the product of genotype freq and fitness

|                    | $A_1A_1$   | $A_1A_2$   | $A_2A_2$   | Total |
| :-------: | : ------ : | :-------: | :-------: | :-------: |
| Initial frq           | $p^2$    | $2pq$    | $q^2$     |  $1$        |
| Coefficient of selection | $0$   | $0$      | $s$       |             |
| Fitness ( $w$ )   |  $1$        |  $1$        |  $1-s$   |            |
| Gametic contribution | $p^2$   | $2pq$    | $q^2(1-s)$    |  $1-sq^2$ |



--

### After one generation of selection against recessive

Note that there has been a proportionate loss of $sq^2$ due to the selection.

The $A_1$ allele freq, 


\begin{align*}
p_1 & = \frac{(2p^2 + 2pq)/2}{p^2 + 2pq + q^2(1-s)} = \frac{p(p+q)}{1 -sq^2}\\
& = \frac{p}{1- sq^2} \\
\end{align*}

---

# Selection

### After one generation of selection against recessive


\begin{align*}
p_1  = \frac{p}{1- sq^2} \\
\end{align*}

### Change in allele freq after one generation

\begin{align*}
\Delta p & = p_1 - p_0  = \frac{p}{1- sq^2} - p\\
& = \frac{p-p(1-sq^2)}{1-sq^2} \\
& = \frac{spq^2}{1 - sq^2}
\end{align*}

--

Then,

\begin{align*}
\Delta q & = -\Delta p = -\frac{spq^2}{1 - sq^2}
\end{align*}

---

# Simulation for selection against recessive

```{r fig.align='center', fig.height=5}
deltap <- function(s, p){
  q <- 1-p
  return((s*p*q^2)/(1 - s*q^2))
}

p <- seq(0, 1, by=0.01)
plot(p, deltap(s=0.4, p), type="l", lwd=3, col="red", xlab="p", ylab="|p1 - p0|")
lines(p, deltap(s=0.2, p), lty=2, lwd=3, col="blue")
```


---

# Effectiveness of Selection

### Initial allele freq

  - Most effective at intermediate freq
  
  - Inefficient when targeted recessive allele is rare

--

### Degree of dominance

- It will become more complex

---

# Degree of dominance

|                    | $A_1A_1$   | $A_1A_2$   | $A_2A_2$   | Total |
| :-------: | : ------ : | :-------: | :-------: | :-------: |
| Initial frq           | $p^2$    | $2pq$    | $q^2$     |  $1$        |
| Fitness ( $w$ ): recessive  |  $1$        |  $1$        |  $1 -s$   |   
| Fitness ( $w$ ): dominance   |  $w_{11}$        |  $w_{12}$        |  $w_{22}$   |            |
| Gametic contribution | $p^2$   | $2pq$    | $q^2(1-s)$    |  $1-sq^2$ |

--

### Recessive:

\begin{align*}
p_1 & = \frac{p^2 + pq}{p^2 + 2pq + q^2(1-s)} = \frac{p}{1- sq^2} \\
\end{align*}

--

### Degree of dominance:

\begin{align*}
p_1 & = \frac{p^2w_{11} + pqw_{12}}{p^2w_{11} + 2pqw_{12} + q^2w_{22}} \\
\end{align*}

---

# Average fitness

\begin{align*}
p_1 & = \frac{p^2w_{11} + pqw_{12}}{p^2w_{11} + 2pqw_{12} + q^2w_{22}} \\
\end{align*}

\begin{align*}
\bar{w} = p^2w_{11} + 2pqw_{12} + q^2w_{22} \\
\end{align*}

- $w_{11}$: genotype specific fitness for $A_1A_1$
- $w_{12}$: genotype specific fitness for $A_1A_2$
- $w_{22}$: genotype specific fitness for $A_2A_2$

\begin{align*}
\Delta p & = \frac{p^2w_{11} + pqw_{12}}{\bar{w}} - p \\
& = \frac{pq(p(w_{11} - w_{12}) + q(w_{12} - w_{22}))}{\bar{w}}
\end{align*}

---

# Practice Problem

In a population of fireflies, allele $A_1$ results in a brighter light. We wish to select for the $A_1$ allele to save money on electricity.

--

36% of the population is $A_1A_1$. We spray all the dimly lit $A_2A_2$ individuals so they are sterile but estimate that we missed 10%, which randomly mate with the others.

--

### Questions:

- What is the fitness of each genotype?

- What type of dominance model is this?

- What is the proportion of fireflies _carrying_ the unwanted, $A_2$ allele after one generation of selection?


---

# Average fitness

- Define each of your variables ( $A_1$, $A_2$, $p$, $q$ )

- Identify % survial/contribution of each genotype

- Calculate the relative fitness ( $w$ ) for each genotype

- Use the general equation with $\bar{w}$

\begin{align*}
p_1 & = \frac{p^2w_{11} + pqw_{12}}{p^2w_{11} + 2pqw_{12} + q^2w_{22}} \\
\end{align*}

- When you get your answer does it make sense?

  - For example, for the favorable allele, i.e., $A_1$ allele, does $p_1$ increase?

---

# Allele freq change over time

Selection for generations to achieve target frequency:

- $p_0$ = current frequency
- $p_t$ = frequency at future time
- $t$ = number of generations

--
----------

### Selection for Dominant allele ( $s=1$ )

It is the same as selection against recessive genotype.

\begin{align*}
p_1 = & \frac{p}{1 - sq^2} \\
= & \frac{1-q}{(1+q)(1-q)} = \frac{1}{1+q}
\end{align*}


--

\begin{align*}
p_2 = & \frac{p_1}{1 - sq^2_1} \\
= & \frac{1-q_1}{(1-q_1^2)} = \frac{1 - (1-p_1)}{1 - (1-p_1)^2} = \frac{\frac{1}{1+q}}{1 - (\frac{q}{1+q})^2}
= \frac{1+q}{1+2q}
\end{align*}


---

# Allele freq change over time

\begin{align*}
p_1  & = \frac{1}{1+q} \\
p_2  & = \frac{1+q}{1+2q} \\
p_3  & = \frac{1+2q}{1+3q} \\
\end{align*}

--

A pattern emerges ...

\begin{align*}
p_t  & = \frac{1+(t-1)q_0}{1+tq_0} \\
q_t  & = 1- p_t = \frac{q_0}{1+tq_0} \\
\end{align*}

--

The generations required to move from $p_0$ to $p_t$ is the same as to move from $q_0$ to $q_t$

\begin{align*}
1 + tq_0 & = \frac{q_0}{q_t} \\
t & = \frac{q_0 - q_t}{q_tq_0} = \frac{1}{q_t} - \frac{1}{q_0}
\end{align*}

---

# Simulation for selection against recessive

\begin{align*}
p_t  & = \frac{1+(t-1)q_0}{1+tq_0} \\
q_t  & = 1- p_t = \frac{q_0}{1+tq_0} \\
\end{align*}

```{r fig.align='center', fig.height=5}
get_pt <- function(p0, t){
  q0 <- 1 - p0
  return(1- q0/(1+t*q0))
  }
```

--

\begin{align*}
t & = \frac{q_0 - q_t}{q_tq_0} = \frac{1}{q_t} - \frac{1}{q_0}
\end{align*}

```{r}

get_t <- function(p0, pt){
  q0 = 1- p0
  qt = 1 - pt
  return(1/qt - 1/q0)
  }
```

---

# Simulation for selection against recessive

```{r fig.align='center', fig.height=5}
t <- 1:50
get_t(p0=0.6, pt=0.9) #7.5
plot(t, get_pt(0.05, t), type="l", lwd=3, col="red", 
     xlab="generations", ylab="p", ylim=c(0, 1))
points(x=get_t(p0=0.05, pt=0.6), y=0.6, col="blue", cex=3, pch=16)
points(x=get_t(p0=0.05, pt=0.9), y=0.9, col="blue", cex=3, pch=16)

```












