---
title: "Chapter 10: Example"
date: "11-19-2019"
author: "Jinliang Yang"
output:
   rmdformats::material:
      highlight: kate
      self_contained: no
      code_folding: hide
---




# A maize experiment

A set of maize hybrids (N=345) crossed from male lines (M1, M2, M3, M4 and M5) and female lines (F1, F2, F3, F4, F5, F6).

The hybrids were planted in four environments, each with three replication.

## Read in data

```{r, eval=TRUE}
corn <- read.csv('https://jyanglab.com/AGRO-931/data/corn.csv')

head(corn)
```

## Phenotypic distribution
```{r}
hist(corn$Yld, main="Yield", xlab="Value", breaks=50, col="#458b74")
```

## Across environments

- To learn about how to do [data visulization](https://www.rstudio.com/resources/cheatsheets/) and other in R.

- Many other [R statistics resources](https://docs.google.com/document/d/1766U2u4LTw19IMSX79XE1on4gRWp6vyN7TGwyOoMOKY/edit)


```{r}
library(ggplot2)

corn$geno <- paste(corn$Male, corn$Female, sep="x")

fsize=18
p1 <- ggplot(data=corn, aes(x=REP, y=Yld, fill= as.factor(REP)) ) +
    geom_violin() +
    facet_wrap(~ ENV) +
    #guides(fill=FALSE) + factor(trait, levels=out$trait)
    labs(y=NULL, fill="Rep") + 
    xlab("") + 
    ylab("") + 
    theme_bw() +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
          axis.text=element_text(size=fsize),
          axis.title=element_text(size=fsize, face="bold"),
          legend.title = element_text(size=fsize, face="bold"),
          legend.text = element_text(size=fsize))
p1
```


# Half-sibs

One parent in common.

```{r}
hs1 <- subset(corn, Male %in% c("M1", "M2"))

fit <- lm(Yld ~ Male + ENV + ENV:REP, data=hs1)
summary(aov(fit))
```

What is the among-families and within-families variance?

The intraclass correlation is `1.9/(1.9 + 3.2)`

### The covariance between two individuals

The covariance between two individuals in the same group is equal to the variance between groups.

Cov(HS) = Var(between)

# Full-sibs

Both parents in common.

```{r}
fs1 <- subset(corn, geno %in% c("M1xF1", "M2xF2"))
fs1

fit <- lm(Yld ~ geno + ENV + ENV:REP, data=fs1)
summary(aov(fit))
```

Again, Cov(FS) = var(between)

# Variance Components

```{r}
corn$REP <- as.factor(corn$REP)

fit <- lm(Yld ~ Male + Female +  ENV + ENV:REP, data=corn)

##Perform simple ANOVA
aovResult <- aov(fit)


##Extract table of mean squares
summary(aovResult)[[1]]

```














