---
title: "Lab1: Basic R for Quant-Gen"
date: "Oct. 9th, 2020"
author: "Jinliang Yang"
output:
   rmdformats::material:
      highlight: kate
      self_contained: no
      code_folding: hide
---


# Data Types in R

For more information, please refer to [Quick-R](https://www.statmethods.net/r-tutorial/index.html).

R has a wide variety of data types including scalars, __vectors__ (numerical, character, logical), matrices, __data frames__, and lists.

### Vectors

```{r}
a <- c(1, 2, 5.3, 6, -2) # numeric vector
b <- c("one","two","three", "four", "five") # character vector
c <- c(TRUE,TRUE,TRUE,FALSE,TRUE) #logical vector
```

Refer to elements of a vector using subscripts.

```{r}
a[c(2,4)] # 2nd and 4th elements of vector
```

### Data Frames

In a data frame, different columns can have different modes (numeric, character, factor, etc.). This is similar to SAS and SPSS datasets.

```{r}
mydata <- data.frame(b, a, c)
names(mydata) <- c("ID","Value","Passed") # variable names
mydata
```


# Operators in R

### Arithmetic Operators include:

| Operator |	Description |
|--------- | -------------|
| +	| addition |
| -	| subtraction |
| *	| multiplication |
| /	| division |
|^ or ** |	exponentiation |

### Logical Operators include:

| Operator | 	Description | 
|--------- | -------------|
|  >	|  greater than | 
|  >=	|  greater than or equal to | 
|  ==	|  exactly equal to | 
|  !=	| not equal to | 


# R for Loop
Loops are used in programming to repeat a specific block of code.

### Syntax of for loop

```{bash, eval=FALSE}
for (val in sequence)
{
statement
}
```

### Example: for loop

Below is an example to count the number of even numbers in a vector.

```{r}
x <- c(2,5,3,9,8,11,6)
count <- 0
for (i in 1:length(x)) {
  if(x[i] %% 2 == 0) {
    count = count + 1
    }
}
print(count)
```


# The Central Limit Theorem (CLT)

- The CLT states that the sums of a set of random variables $(X_1, X_2, X_3, ..., X_n)$ is normally
distributed no matter the distribution the individual X's were sampled from, 
as long as they were sampled from __identical distributions__.

### A simulation experiment

\begin{align*}
Y_{i} = \sum\limits_{j=1}^{j=m} X_{ij} \alpha_{j} + \epsilon_i
\end{align*}


- For a given individual ( $i=1$ ) with a number of loci ( $m=1,000$ )
- Each allele is $X_j \in (A, a)$ , with the probability of $p$ or $1-p$
- The effect of $j$th allele ( $\alpha_j$ ) can be samples from any distribution (e.g., _uniform distribution_)

According to the CLT, if $m$ is __sufficiently large__, the sum is normally distributed.

### Simulate one dividual

Simulate an individual's phenotypic value. In this individual, the phenotype is determined by `m` number of markers with marker freq = 0.5. The markers' effects ( $\alpha$ ) are randomly draws from a uniform distribution.

```{r}
m <- 1000
## for each allele, the chance of A or a is equal to 0.5
x <- rbinom(n=m, size=1, prob=0.5)
## sample effect from a uniform distribution:
a <- runif(n=m)
y <- sum(x*a) + 0
y
```

Apply the above procedure to a population composed of `n` individuals.

```{r}
set.seed(1234) # seed for random number generator
m <- 1000
n <- 2000 # simulate a population of 2,000 individuals
out <- c() # create an empty vector
for(i in 1:n){ #<<
  x <- rbinom(m, size=1, prob=0.5) ## for each allele, the chance of A = 0.5
  a <- runif(m) ## sample effect from a uniform distribution:
  y <- sum(x*a)
  out <- c(out, y)
} 

```

### Plot the result
```{r}
#shapiro.test(out) # W = 0.99928, p-value = 0.6622
hist(out, breaks=50, col="#b8860b", main="Phenotype Distribution", xlab="")
```

### Determine what does the __sufficiently large__ mean?

Pack the abvoe simulation procedure into an R function:
```{r}
sim_clt <- function(m=1, n=2000){
  # m: number of markers, m=1
  # n: number of individuals, n=2000
  out <- c() # create an empty vector
  for(i in 1:n){ #<<
    x <- rbinom(m, size=1, prob=0.5) ## for each allele, the chance of A = 0.5
    a <- runif(m) ## sample effect from a uniform distribution:
    y <- sum(x*a)
    out <- c(out, y)
  } 
  # output p.value
  return(shapiro.test(out)$p.value)
}
```

Then apply the function using an R for loop:

We will run a sequene of number of markers from 10 to 1000, with the increment of 10. 

```{r}
set.seed(12345)
pval <- c() # create an empty vector as the output
# 
num <- seq(from =10, to =1000, by=10) # 
for(i in num){
  # here we apply the function for the situation with i markers
  tem <- sim_clt(m=i, n=2000)
  pval <- c(pval, tem)
}

```

Again, let's plot the result!

```{r}
plot(x=num, y=-log10(pval), col="#b8860b", main="Phenotype Distribution", xlab="")
abline(h=-log10(0.01))
```


