---
title: "Changes in allele frequency"
author: "Jinliang Yang"
date: "Aug. 29, 2022"
output:
  xaringan::moon_reader: 
    css: ["default", "../../asset/unl-150.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Factors lead  to changes of allele freq

### Systematic processes

Predictable both in amount and in direction

- __Migration__

- __Mutation__

- __Selection__

--

### Stochastic process

Arises in small populations from the effect of sampling.

- Predictable in amount but not in direction

---

# Migration

- Rate of change depends on rate of migration and difference in allele frequency between original pop and immigrants

- In agriculture, can think of cross-breeding as having a similar effect as migration

--

### In mixed population after migration

- $m$ = proportion of new immigrants (each generation)
- $q_m$ = allele freq in immigrant population
- $q_0$ = allele freq in original pop

--

\begin{align*}
q_1 & = mq_m + (1-m)q_0 \\
 & = m(q_m - q_0) + q_0 \\
\end{align*}

--

Change in allele freq:

\begin{align*}
\Delta q & = q_1 - q_0 = m(q_m - q_0)
\end{align*}

---

# Mutation

- To affect allele frequencies, a mutation has to be maintained!

- New mutation = heterozygote (most commonly)
  - 50% chance of being passed on 
  - After $n$ generations, the probability of the allele remains

\begin{align*}
(1 - \frac{1}{2})^n
\end{align*}

--

```{r echo=FALSE, fig.height=5}
n <- 1:10
plot(x=n, y=(1-1/2)^n, type="l", lwd=3)
```

---

# Mutation

### What is the true rate of mutation ( $\mu$ )? 

- Depends on organism and locus

--

- Human:
  - Point mutation (2.5 to 1 $\times 10^{-8}$ mutations/locus/generation)
    > Neel et al 1986; Nachman and Crowell 2000

- Viruses
  - $1 \times 10^{-3}$ to $1 \times 10^{-5}$ per base/generation
  
  - SARS-CoV-2: spontaneous mutation rate of $1.3 \times 10^{-6}$ (this study found spike protein accumulated five times more mutations => consistent with the action of positive selection)
   > Amicone et al 2022


---

# Recurrent Mutation

- Forward mutation rate __ $\mu$__ from $A_1$ -> $A_2$
- Backward mutation rate __ $\nu$__ from $A_1$ <- $A_2$

- Given the initial allele freq $p_0$ for $A_1$ and $q_0$ for $A_2$

--

- Change $\Delta q = \mu p_0 - \nu q_0$

--

In equilibrium,

\begin{align*}
&\Delta q = \mu p_0 - \nu q_0 = 0 \\
& \mu p_0 = \nu q_0 \\
\end{align*}

or 

\begin{align*}
& \mu (1- q) = \nu q \\
& q = \frac{\mu}{\nu + \mu} \\
\end{align*}

---

# Mutation

- #### Mutation alone produces slow change in allele frequency

- #### But, mutation may be important in some small pops (later)

- #### And, mutation is the ultimate source of genetic variation

---

# Selection

Selection occurs due to differences in viability and fertility (natural or artificial)

--

If fitness varies by genotype, selection will occur

- __Directional__: favors phenotypes at one extreme of distribution range

- __Stabilizing__: favors intermediate phenotypes

- __Disruptive__: favors phenotypes at both extremes

---

# Selection (and Fitness)

The contribution of offspring to the next generation is called the __fitness__ of the individual, or _selective value_.

- Selection coefficient, __ $s$__, is the proportionate __REDUCTION__ in gametic contribution of a particular genotype _compared to the standard genotype_

- Relative fitness $w = 1 -s$
  - Most fit is set to 1
  

- e.g., $AA$, $Aa$, and $aa$ have survival of 0.75, 0.75, and 0.5, respectively

--

  - Then, the relative fitness is 1.0, 1.0, and 0.67 (0.5/0.75)

--

----------------------

- Dominance needs to be accounted for when quantifying allele frequency changes with selection.

- The degree of dominance influences the relative fitness of alleles. 

- Here dominances is with respect to fitness only.

---

# Selection with dominance

In this example, __ $A_1$ is the most favorable allele__, conferring the greatest degree of fitness

| Degree of dominance | $A_2A_2$   | $A_1A_2$   | $A_1A_1$   |
| :-------: | : ------ : | :-------: | :-------: |
| Additive           | $1 - s$    | $1 - s/2$    | $1$     | 
| Partial dominance   |  $1 -s$        |  $1 - hs$        |  $1$   | 
| Complete dominance | $1 -s$   | $1$    | $1$    |
| Overdominance      | $1 -s_2$ | $1$ | $1-s_1$ | 

--

Here, $h$ is the heterozygous effect (or level of dominance) from the fitness of the heterozygote relative to the difference between homozygotes.

---

# Selection against the recessive allele

- Gametic contribution is the product of genotype freq and fitness

|                    | $A_1A_1$   | $A_1A_2$   | $A_2A_2$   |
| :-------: | : ------ : | :-------: | :-------: |
| Initial frq           | $p^2$    | $2pq$    | $q^2$     | 
| Coefficient of selection | $0$   | $0$      | $s$       |
| Fitness ( $w$ )   |  $1$        |  $1$        |  $1-s$   | 
| Gametic contribution | $p^2$   | $2pq$    | $q^2(1-s)$    |

---

# Selection against the recessive allele

- Gametic contribution is the product of genotype freq and fitness

|                    | $A_1A_1$   | $A_1A_2$   | $A_2A_2$   | Total |
| :-------: | : ------ : | :-------: | :-------: | :-------: |
| Initial frq           | $p^2$    | $2pq$    | $q^2$     |  $1$        |
| Coefficient of selection | $0$   | $0$      | $s$       |             |
| Fitness ( $w$ )   |  $1$        |  $1$        |  $1-s$   |            |
| Gametic contribution | $p^2$   | $2pq$    | $q^2(1-s)$    |  $1-sq^2$ |



--

### After one generation of selection against recessive

Note that there has been a proportionate loss of $sq^2$ due to the selection.

The $A_1$ allele freq, 


\begin{align*}
p_1 & = \frac{(2p^2 + 2pq)/2}{p^2 + 2pq + q^2(1-s)} = \frac{p(p+q)}{1 -sq^2}\\
& = \frac{p}{1- sq^2} \\
\end{align*}

---

# Selection

### After one generation of selection against recessive


\begin{align*}
p_1  = \frac{p}{1- sq^2} \\
\end{align*}

### Change in allele freq after one generation

\begin{align*}
\Delta p & = p_1 - p_0  = \frac{p}{1- sq^2} - p\\
& = \frac{p-p(1-sq^2)}{1-sq^2} \\
& = \frac{spq^2}{1 - sq^2}
\end{align*}

--

Then,

\begin{align*}
\Delta q & = -\Delta p = -\frac{spq^2}{1 - sq^2}
\end{align*}

---

# Simulation for selection against recessive

```{r fig.align='center', fig.height=5}
deltap <- function(s, p){
  q <- 1-p
  return((s*p*q^2)/(1 - s*q^2))
}

p <- seq(0, 1, by=0.01)
plot(p, deltap(s=0.4, p), type="l", lwd=3, col="red", xlab="p", ylab="|p1 - p0|")
lines(p, deltap(s=0.2, p), lty=2, lwd=3, col="blue")
```











