---
title: "Examples for heritability estimation"
author: "Jinliang Yang"
date: "April 10, 2024"
output:
  xaringan::moon_reader:
    css: ["default", "../../asset/unl-150.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
---

# A sib analysis

### Danish Landrace pigs
> Data based on the $n=198$ Danish pig progeny testing records

--

- 2 sires each mated to 34 dams ( $s=2, d=34$ )

- Recorded body weight of 2 male and 2 female offspring $n=4$ from each dam

---
# Danish landrace pigs

```{r}
p <- read.csv('https://jyanglab.com/slides/swine_sib.csv')
dim(p)
table(p$sire)
```

```{r}
table(p$dam)
```

---
# ANOVA

```{r}
fit <- lm(weight ~ sire + dam, data=p)
summary(aov(fit))
```

#### __ANOVA__ table for body length:

| Source         |  df    |    Mean Sq    |  Expectation of MS |
| :------------: | :-------: | :-------:     | :-------: |
| Between Sires  | 1     |    $MS_s=6331$     | $= \sigma_w^2 + n\sigma_d^2 + dn\sigma_s^2$ |
| Between Dams   | 33      |    $MS_d=734$     | $= \sigma_w^2 + n\sigma_d^2$ | 
| Within progenies | 163    |    $MS_w=352$     | $= \sigma_w^2$ |

---

# Components of variance

#### __ANOVA__ table for body length:

| Source         |  df    |    Mean Sq    |  Expectation of MS |
| :------------: | :-------: | :-------:     | :-------: |
| Between Sires  | 1     |    $MS_s=6331$     | $= \sigma_w^2 + n\sigma_d^2 + dn\sigma_s^2$ |
| Between Dams   | 33      |    $MS_d=734$     | $= \sigma_w^2 + n\sigma_d^2$ | 
| Within progenies | 163    |    $MS_w=352$     | $= \sigma_w^2$ |


Because $d=34$ and $n=4$ as we designed

\begin{align*}
& \sigma_w^2 = 352 \\
\end{align*}

--

\begin{align*}
& \sigma_d^2 = (734 - 352)/4 = 95.5 \\
\end{align*}

--


\begin{align*}
& \sigma_s^2 = (6331 - 734)/(34 \times 4) = 41.15 \\
\end{align*}



---

# Summary of the variance components

| Observational  | Covariance and causal components  |  Estimated values  |
| :------------: | :-------: |  | 
| Sires | $\sigma_s^2 = Cov(HS)$  | $=\frac{1}{4}\sigma_A^2$    |
| Dams  | $\sigma_d^2 = Cov(FS) - Cov(HS)$ | $=\frac{1}{4}\sigma_A^2 + \frac{1}{4}\sigma_D^2$ | 
| Progeny | $\sigma_w^2 = V_P - Cov(FS)$  | $= \frac{1}{2}\sigma_A^2 +\frac{3}{4}\sigma_D^2$ |
| Total | $\sigma_T^2 = V_P = \sigma_s^2 + \sigma_d^2 + \sigma_w^2$ | $=\sigma_A^2 + \sigma_D^2 + \sigma_E^2$ |
| Sires + Dams   | $\sigma_s^2 + \sigma_d^2 = Cov(FS)$       | $=\frac{1}{2}\sigma_A^2 + \frac{1}{4}\sigma_D^2$ |

--

\begin{align*}
V_P = \sigma_T^2 & = \sigma^2_s + \sigma^2_d + \sigma^2_w \\
& = 41.15 + 95.5 + 352 = 488.65 \\
\end{align*}

--

### Estimates of heritability from Sire-component

\begin{align*}
h^2 = \frac{V_A}{V_P} = 4 \times \frac{\sigma^2_s}{\sigma_T^2} = 4 \times \frac{41.15}{488.65} = 0.34
\end{align*}

---

# Summary of the variance components

| Observational  | Covariance and causal components  |  Estimated values  |
| :------------: | :-------: |  | 
| Sires | $\sigma_s^2 = Cov(HS)$  | $=\frac{1}{4}\sigma_A^2$    |
| Dams  | $\sigma_d^2 = Cov(FS) - Cov(HS)$ | $=\frac{1}{4}\sigma_A^2 + \frac{1}{4}\sigma_D^2$ | 
| Progeny | $\sigma_w^2 = V_P - Cov(FS)$  | $= \frac{1}{2}\sigma_A^2 +\frac{3}{4}\sigma_D^2$ |
| Total | $\sigma_T^2 = V_P = \sigma_s^2 + \sigma_d^2 + \sigma_w^2$ | $=\sigma_A^2 + \sigma_D^2 + \sigma_E^2$ |
| Sires + Dams   | $\sigma_s^2 + \sigma_d^2 = Cov(FS)$       | $=\frac{1}{2}\sigma_A^2 + \frac{1}{4}\sigma_D^2$ |

--

\begin{align*}
V_P = \sigma_T^2 & = \sigma^2_s + \sigma^2_d + \sigma^2_w \\
& = 41.15 + 95.5 + 352 = 488.65 \\
\end{align*}

--

### Estimates of heritability from Dam-component

\begin{align*}
H^2 = \frac{V_A + V_D}{V_P} = 4 \times \frac{\sigma^2_d}{\sigma_T^2} = 4 \times \frac{95.5}{488.65} = 0.78
\end{align*}

---

# Calculate sib correlations

### Half-sibs

\begin{align*}
t_{HS} & = \frac{Cov(HS)}{V_P} = \frac{1/4V_A}{V_P} = 1/4 h^2\\
  & = \frac{\sigma^2_s}{\sigma_T^2}  =  \frac{41.15}{488.65} = 0.08 \\
\end{align*}

--

### Full-sibs

\begin{align*}
t_{FS} & = \frac{Cov(FS)}{V_P} = \frac{1/2V_A + 1/4V_D}{V_P}\\
&= \frac{\sigma^2_s + \sigma^2_d}{\sigma_T^2}  =  \frac{41.15 + 95.5}{488.65} = 0.28 \\
\end{align*}


---

# Cross-factorial design: a maize example

### North-Carolina Design II

- A set of maize hybrids crossed from male lines (M1, M2, M3, M4 and M5) and female lines (F1, F2, F3, F4, F5, F6).
  - $m=5$ and $f=6$

- The hybrids were planted in four environments, each with three replications.

--

### Read in data

```{r, eval=TRUE}
corn <- read.csv('https://jyanglab.com/AGRO-931/data/corn.csv')

head(corn, 3)
```

---

### Phenotypic distributions across environments

```{r, eval=FALSE}
library(ggplot2)

corn$geno <- paste(corn$Male, corn$Female, sep="x")

fsize=18
p1 <- ggplot(data=corn, aes(x=REP, y=Yld, fill= as.factor(REP)) ) +
    geom_violin() +
    facet_wrap(~ ENV) +
    #guides(fill=FALSE) + factor(trait, levels=out$trait)
    labs(y=NULL, fill="Rep") + 
    xlab("") + 
    ylab("") + 
    theme_bw() +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
          axis.text=element_text(size=fsize),
          axis.title=element_text(size=fsize, face="bold"),
          legend.title = element_text(size=fsize, face="bold"),
          legend.text = element_text(size=fsize))
p1
```

---

### Phenotypic distributions across environments

```{r echo=FALSE, fig.height=5}
library(ggplot2)

corn$geno <- paste(corn$Male, corn$Female, sep="x")

fsize=18
p1 <- ggplot(data=corn, aes(x=REP, y=Yld, fill= as.factor(REP)) ) +
    geom_violin() +
    facet_wrap(~ ENV) +
    #guides(fill=FALSE) + factor(trait, levels=out$trait)
    labs(y=NULL, fill="Rep") + 
    xlab("") + 
    ylab("") + 
    theme_bw() +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
          axis.text=element_text(size=fsize),
          axis.title=element_text(size=fsize, face="bold"),
          legend.title = element_text(size=fsize, face="bold"),
          legend.text = element_text(size=fsize))
p1
```

---

# Cross-factorial design

\begin{align*}
p_{ijk} = \mu + m_i + f_j + I_{ij} + e_{ijk}
\end{align*}

- here $I_{ij}$ indicate male and female interaction 
  - allows for non-additive effect of interactions between alleles that were inherited from both the male and female

```{r}
fit <- lm(Yld ~ Male + Female + Male:Female + ENV + REP, data=corn)
summary(aov(fit))
```

---


# ANOVA table for a factorial design

| Source         |  df    |    Mean Sq    |  Expectation of MS |
| :------------: | :-------: | :-------:     | :-------: |
| Env  | $v-1$     |      | |
| Reps in env  | $v(r-1)$     |    | |
| Between Sires (male)  | $v(m-1)$     |    $MS_f$     | $= \sigma_e^2 + r\sigma_i^2 + rf\sigma_m^2$ |
| Between Dams (female)   | $v(f-1)$      |    $MS_m$     | $= \sigma_e^2 + r\sigma_i^2 + rm\sigma_f^2$ | 
| Interaction   | $v(s-1)(f-1)$      |    $MS_i$     | $= \sigma_e^2 + r\sigma_i^2$ | 
| Plots within rep (error) | $v(mf-1)(r-1)$    |    $MS_e$     | $= \sigma_e^2$ |

- $v$ is the number of environments
- $r$ is the number of reps per environment
- $f$ is the number of female lines
- $m$ is the number of male lines

---


# In terms of causal variances

| Source         |  df    |    Mean Sq    |  Expectation of MS |
| :------------: | :-------: | :-------:     | :-------: |
| Between Sires (male)  | $v(m-1)$     |    $MS_f$     | $= \sigma_e^2 + r\sigma_i^2 + rf\sigma_m^2$ |
| Between Dams (female)   | $v(f-1)$      |    $MS_m$     | $= \sigma_e^2 + r\sigma_i^2 + rm\sigma_f^2$ | 
| Interaction   | $v(s-1)(f-1)$      |    $MS_i$     | $= \sigma_e^2 + r\sigma_i^2$ | 
| Plots within rep (error) | $v(mf-1)(r-1)$    |    $MS_e$     | $= \sigma_e^2$ |



#### Estimate from ANOVA

| Observational component        |  Value     |    Resemblance of relatives    |  
| :------------: | :-------: | :--------------------:     | :-------: |
| Male variance  | $\sigma_m^2=\frac{1}{rf}(MS_f - MS_i)$     |   $= Cov(MHS)$   |
| Female variance  | $\sigma_f^2=\frac{1}{rm}(MS_m - MS_i)$     |  $= Cov(FHS)$  | 
| Interaction   | $\sigma_i^2=\frac{1}{r}(MS_i - MS_e)$      |    $= Cov(FS) - Cov(MHS) - Cov(FHS)$    | 
| Error variance | $MS_e$    |   $= V_P - Cov(FS)$   | 

---

# Back to the maize data


```{r}
summary(aov(fit))
```

| Source         |  df    |    Mean Sq    |  Expectation of MS |
| :------------: | :-------: | :-------:     | :-------: |
| Env  | $v-1$     |      | |
| Reps in env  | $v(r-1)$     |    | |
| Between Sires (male)  | $v(m-1)$     |    $MS_f=14.96$     | $= \sigma_e^2 + r\sigma_i^2 + rf\sigma_m^2$ |
| Between Dams (female)   | $v(f-1)$      |    $MS_m=16.25$     | $= \sigma_e^2 + r\sigma_i^2 + rm\sigma_f^2$ | 
| Interaction   | $v(s-1)(f-1)$      |    $MS_i=6.60$     | $= \sigma_e^2 + r\sigma_i^2$ | 
| Plots within rep (error) | $v(mf-1)(r-1)$    |    $MS_e=4.15$     | $= \sigma_e^2$ |


---

# Estimate from ANOVA



| Observational component        |  Value     |    Resemblance of relatives    |  
| :------------: | :-------: | :--------------------:     | :-------: |
| Male variance  | $\sigma_m^2=\frac{1}{rf}(MS_f - MS_i)$     |   $= Cov(MHS)$   |
| Female variance  | $\sigma_f^2=\frac{1}{rm}(MS_m - MS_i)$     |  $= Cov(FHS)$  | 
| Interaction   | $\sigma_i^2=\frac{1}{r}(MS_i - MS_e)$      |    $= Cov(FS) - Cov(MHS) - Cov(FHS)$    | 
| Error variance | $MS_e$    |   $= V_P - Cov(FS)$   | 

- $v=4, r=3, m=5, f=6$
- $MS_f=14.96, MS_m=16.25, MS_i=6.60, MS_e=4.15$

\begin{align*}
& \sigma_m^2 = \frac{1}{3 \times 6}(14.96 - 6.60) = 0.46\\
\end{align*}

--

\begin{align*}
& \sigma_f^2 = \frac{1}{3 \times 5}(16.25 - 6.60) = 0.64\\
\end{align*}


--


\begin{align*}
& \sigma_i^2 = \frac{1}{3}(6.60 - 4.15) = 0.81\\
\end{align*}


---
# Heritablity estimation

\begin{align*}
p_{ijk} = \mu + m_i + f_j + I_{ij} + e_{ijk}
\end{align*}

--

\begin{align*}
V_P & = V_m + V_f +V_I + V_e \\
& = \sigma_m^2 + \sigma_f^2 + \sigma_i^2 + \sigma_e^2 \\
&= 0.46+ 0.64 + 0.81 + 4.15 = 6.06 \\
\end{align*}

--

----------

Heritability among half-sib families can be estimed either the male or female half-sib variance component or the everage of the two.

\begin{align*}
h^2_{f} & = 4 \times \frac{\sigma_f^2}{\sigma_T^2} \\
& = 4 \times \frac{0.64}{6.06} = 0.42\\
\end{align*}

--

\begin{align*}
h^2_{m} & = 4 \times \frac{\sigma_m^2}{\sigma_T^2} \\
& = 4 \times \frac{0.46}{6.06} = 0.30\\
\end{align*}






