---
title: "Chapter 9: Example"
date: "10-26-2018"
author: "Jinliang Yang"
output:
   rmdformats::material:
      highlight: kate
      self_contained: no
      code_folding: hide
---




# A sib analysis

A set of maize hybrids (N=345) crossed from male lines (M1, M2, M3, M4 and M5) and female lines (F1, F2, F3, F4, F5, F6).

The hybrids were planted in four environments, each with three replication.

## Read in data

```{r, eval=TRUE}
pigData <- read.csv('https://jyanglab.com/AGRO-931/data/pigsibs.csv')

p <- pigData
table(p)

length(table(p$Sire))

p <- pigData[, c(1,3,4,5)]
names(p)[3:4] <- c("dam", "sire")
p$dam <- paste0("d", p$dam)
p$sire <- paste0("s", p$sire)
p$cross <- paste(p$sire, p$dam, sep="x")


p$kid <- 1
c <- unique(p$cross)
df <- data.frame()
for(i in 1:length(c)){
  sub <- subset(p, cross %in% c[i])
  if(nrow(sub) > 4){
    sub <- sub[1:4, ]
  }
  sub$kid <- 1:nrow(sub)
  df <- rbind(df, sub)
}

names(df)[2] <- "weight"

write.table(df, "swine_sib.csv", sep=",", row.names = FALSE, quote=FALSE)

fit <- lm(weight ~ sire + dam, data=df)
summary(aov(fit))

```

```{r}
vw <- 352
n=4 # number of kids per full-sib family
vd <- (734 - vw)/n 
d=33
vs <- (6331 - 734)/(d*n)

vp <- vd + vs + vw
```

$\sigma_s^2 = Cov(HS) = 1/4 \sigma_A^2$

```{r}
va <- 4*vs
```

$h^2 = V_A/V_P$

```{r}
va/vp
```

Vd

$\sigma_d^2 = Cov(FS) - Cov(HS) = 1/4 \sigma_A^2 + 1/4 \sigma_D^2$

```{r}
Vd = 4*vd  - va
```


## Phenotypic distribution
```{r}
hist(corn$Yld, main="Yield", xlab="Value", breaks=50, col="#458b74")
```

## Across environments

- To learn about how to do [data visulization](https://www.rstudio.com/resources/cheatsheets/) and other in R.

- Many other [R statistics resources](https://docs.google.com/document/d/1766U2u4LTw19IMSX79XE1on4gRWp6vyN7TGwyOoMOKY/edit)


```{r}
library(ggplot2)

corn$geno <- paste(corn$Male, corn$Female, sep="x")

fsize=18
p1 <- ggplot(data=corn, aes(x=REP, y=Yld, fill= as.factor(REP)) ) +
    geom_violin() +
    facet_wrap(~ ENV) +
    #guides(fill=FALSE) + factor(trait, levels=out$trait)
    labs(y=NULL, fill="Rep") + 
    xlab("") + 
    ylab("") + 
    theme_bw() +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
          axis.text=element_text(size=fsize),
          axis.title=element_text(size=fsize, face="bold"),
          legend.title = element_text(size=fsize, face="bold"),
          legend.text = element_text(size=fsize))
p1
```


# Half-sibs

One parent in common.

```{r}
hs1 <- subset(corn, Male %in% c("M1", "M2"))

fit <- lm(Yld ~ Male + ENV + ENV:REP, data=hs1)
summary(aov(fit))
```

What is the among-families and within-families variance?

The intraclass correlation is `1.9/(1.9 + 3.2)`

### The covariance between two individuals

The covariance between two individuals in the same group is equal to the variance between groups.

Cov(HS) = Var(between)

# Full-sibs

Both parents in common.

```{r}
fs1 <- subset(corn, geno %in% c("M1xF1", "M2xF2"))
fs1

fit <- lm(Yld ~ geno + ENV + ENV:REP, data=fs1)
summary(aov(fit))
```

Again, Cov(FS) = var(between)

# Variance Components

```{r}
corn$REP <- as.factor(corn$REP)

fit <- lm(Yld ~ ENV + ENV:REP + Male + Female + Male*Female + Male*ENV + Female*ENV + Male*Female*ENV, data=corn)

##Perform simple ANOVA
aovResult <- aov(fit)


##Extract table of mean squares
summary(aovResult)[[1]]

```














